---
title: "HW7b - Probability and the Normal Distribution"
---

In this assignment, you'll build on your z-score skills to answer probability questions using the normal distribution. You'll return to the White House salaries data and use it to answer the kinds of "what are the chances?" questions that come up in journalism all the time. This sets the stage for sampling, which we'll cover next.

### Task 1: Load the tidyverse and the data (1 point)

Load the tidyverse and the White House salaries data from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv into a dataframe called `wh_salaries`

```{r}
library(tidyverse)
wh_salaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv")
```

### Task 2: Filter and summarize (2 points)

You may remember from HW3 that some White House staffers have a salary of $0. Let's filter those out and focus on paid employees, saving the result to a new dataframe called `wh_paid`. You need to calculate the mean, median and standard deviation of the salary column using `wh_paid`:

```{r}
wh_paid <- wh_salaries |>
  filter(SALARY > 0) |>
  
mutate (
mean_salaries = mean(SALARY),
median_salaries = median(SALARY),
sd_salaries = sd(SALARY) 
)

```

Now create a histogram of the paid salaries with a vertical line for the mean and two more for ±1 standard deviation, similar to what you did in HW6 with Spotify data.

```{r}
wh_paid |>
  ggplot() +
  geom_histogram(aes(x=SALARY), binwidth=5000) +
geom_vline(aes(xintercept = mean(SALARY)), color = "red", linetype = "dashed", size = 1) + 
geom_vline(aes(xintercept = mean(SALARY) - sd(SALARY)), color = "blue", linetype = "dashed", size = 1) +
geom_vline(aes(xintercept = mean(SALARY) + sd(SALARY)), color = "blue", linetype = "dashed", size = 1)
```

**Describe the shape of the distribution. Is it roughly normal? Skewed? Where do most salaries cluster?**

This data is skewed to the right. It has some high points later on, but most of the trend sees the data spike early and then decline later. Most salaries lie to the left of the mean, and you can see a cluster around the 100,000 benchmark. 

### Task 3: The Empirical Rule (3 points)

The empirical rule (also called the 68-95-99.7 rule) says that for a normal distribution:
- About 68% of data falls within 1 standard deviation of the mean
- About 95% falls within 2 standard deviations
- About 99.7% falls within 3 standard deviations

Let's test this. We'll calculate the salary ranges for 1, 2, and 3 standard deviations from the mean, then count how many employees actually fall in each range.

```{r}
mean_sal <- mean(wh_paid$SALARY)
sd_sal <- sd(wh_paid$SALARY)
total_n <- nrow(wh_paid)

empirical <- tibble(
  range = c("Within 1 SD", "Within 2 SD", "Within 3 SD"),
  lower = c(mean_sal - sd_sal, mean_sal - 2 * sd_sal, mean_sal - 3 * sd_sal),
  upper = c(mean_sal + sd_sal, mean_sal + 2 * sd_sal, mean_sal + 3 * sd_sal),
  expected_pct = c(68, 95, 99.7),
  actual_count = c(
    sum(wh_paid$SALARY >= (mean_sal - sd_sal) & wh_paid$SALARY <= (mean_sal + sd_sal)),
    sum(wh_paid$SALARY >= (mean_sal - 2 * sd_sal) & wh_paid$SALARY <= (mean_sal + 2 * sd_sal)),
    sum(wh_paid$SALARY >= (mean_sal - 3 * sd_sal) & wh_paid$SALARY <= (mean_sal + 3 * sd_sal))
  ),
  actual_pct = round((actual_count / total_n) * 100, 1)
)

empirical
```

**How closely does the data follow the empirical rule? What does any deviation tell you about the distribution?**

The data is a little under the expected percentages for the first 2 windows, but eclipses the third expected percentage. However, all three percentages are close enough that you could consider this data to be close to empirical. The deviation tells me that the distribution is not normal, and has some kind of skew to alter the empirical rule.

### Task 4: From z-scores to probability with pnorm() (4 points)

Remember the car thefts z-score activity from class? Well, I want to return to that briefly, introduce a new tool called pnorm() and then apply it here. So open that CarTheftZScores notebook and, at the bottom, copy and paste in the contents here: https://gist.githubusercontent.com/dwillis/25102889f576b1707576b85e0be0c353/raw/1ca2be9636497cb5293ff7b1f4e6ed8e27a9e049/car_theft_pnorm.rmd

Re-run all of the blocks in CarTheftZScores notebook, including the code you just pasted in. What's happening there is you are using a function called `pnorm()` to find the probability of a value being less than or equal to a certain z-score. The basic formula to find that probability is:

```
1 - pnorm(z_score)
```

Where `z_score` is one row's z-score. Now apply the same approach to salaries.

Your editor asks: **"What's the probability that a randomly selected White House employee makes more than $150,000?"**

To answer this:
1. Calculate the z-score for $150,000
2. Use `pnorm()` to find the probability. `pnorm()` gives you the area under the normal curve to the LEFT of a z-score — that is, the probability of being BELOW that value.
3. Subtract from 1 to get the probability of being ABOVE $150,000
4. Check your answer against the actual data

```{r}

salary <- 150000
z_score <- (salary - mean_sal) / sd_sal

prob_less <- pnorm(z_score)
prob_more <- 1 - prob_less
prob_more

# Check against actual data
actual_prob <- sum(wh_paid$SALARY > 150000) / total_n
actual_prob
```

**Write a sentence describing this probability in plain English. How does the normal approximation (prob_less) compare to the actual percentage?**

The normal approximation of 16.9% is lower than the actual percentage of 18.2%. Therefore, the actual probability that a randomly selected White House employee makes over $150,000 is is more likely than what was expected from a normal distribution.

### Task 5: Probability between two values (3 points)

Now your editor asks: **"What percentage of White House employees make between $80,000 and $120,000?"**

You'll need to calculate z-scores for those two figures, use `pnorm()` for each, and subtract to find the probability between them. Update the REPLACE_ME figures with the appropriate numbers.

```{r}

z_lower <- (80000 - mean_sal) / sd_sal
z_upper <- (120000 - mean_sal) / sd_sal

prob_between <- pnorm(z_upper) - pnorm(z_lower)
prob_between

# Verify with actual data
actual_between <- sum(wh_paid$SALARY >= 80000 & wh_paid$SALARY <= 120000) / total_n
actual_between

```

**Interpret this result. What does this "middle range" tell us about the White House salary structure?**

The expected middle range is 36.6%, while the actual middle range is 32.6%. This tells me that the White House salary structure has less people between $80000 and $120000 than a normal distribution would have, which could imply that there are more people with an extremely high or extremely low salary than there would be normally. 

### Task 6: Expected vs. observed (3 points)

Let's compare what the normal distribution *predicts* to what we actually *see*. This idea — comparing expected to observed — is one you'll use again later in the course.

Create these salary bins and calculate the expected count (from the normal distribution) and the observed count (from the actual data) for each:

```{r}
bins <- tibble(
  category = c("Below $60k", "$60k-$90k", "$90k-$120k", "$120k-$150k", "Above $150k"),
  lower = c(-Inf, 60000, 90000, 120000, 150000),
  upper = c(60000, 90000, 120000, 150000, Inf)
) |>
  mutate(
    expected_pct = pnorm(upper, mean = mean_sal, sd = sd_sal) - pnorm(lower, mean = mean_sal, sd = sd_sal),
    expected_count = round(expected_pct * total_n, 1),
    observed_count = c(
      sum(wh_paid$SALARY < 60000),
      sum(wh_paid$SALARY >= 60000 & wh_paid$SALARY < 90000),
      sum(wh_paid$SALARY >= 90000 & wh_paid$SALARY < 120000),
      sum(wh_paid$SALARY >= 120000 & wh_paid$SALARY < 150000),
      sum(wh_paid$SALARY >= 150000)
    ),
    difference = observed_count - expected_count
  )

bins |> select(category, expected_count, observed_count, difference)
```

**Where does reality differ most from what the normal distribution predicts? What might explain those differences?**

Reality differs the most in the bottom of the salary distribution. The normal distribution predicts many more people to be below $60,000 and predicts many less people to have salaries between $60k and $90k. This could be because the salary wouldn't be enough to live off of if it was far below $60k, which the normal distribution wouldn't know. What probably happened is all the 'lowest-paid employees' have a higher floor than the normal distribution would expect, meaning there are a lot of people in the second-lowest bin. 

### Task 7: Story pitch (4 points)

Based on your probability analysis, write a headline and lede for a story about White House salaries. Your lede should incorporate at least one probability or percentage you calculated.

Then write 2-3 sentences describing how you would use probability thinking to advance this story — what could you compare these probabilities to, and what additional analysis would help?

HEADLINE: Majority of White House salaries are over $100k

Data shows that over 50% of White House salaries are over 100,000, with 81 people holding salaries over $150k. This figure is much higher than what would be expected in a normal distribution, which calculated an expected number of 66.9 - a 21.1% increase. In addition, only 32.6% of employees have a salary between $80k-$120k, which is also lower than the expected total from a normal distribution. Cross-referencing these two figures, it's revealed that White House salaries see an extreme leap that isn't representative of a normal distribution.

Probability thinking: Probability thinking would advance this story by shifting the focus from a single, static number to a dynamic, risk-adjusted, and context-aware narrative. It replaces oversimplified narratives of "hard work equals high pay" with a more accurate picture that includes luck, base rates, and volatility. I could compare this to other government agencies such as members of Congress and analyze if there are any coorelations between the two in terms of how the salary spread looks.

You could advance this story with probability thinking to see how likely it is for a salary grouping to have this skewed distribution. This could help readers understand that this distribution is not usual, and maybe inspire deeper research into the topic. Additional analysis that could help are comparisons to the probabilities of other government agencies and their salaries, and see if there are other groups that have similarly skewed or unlikely distributions.


### Task 8: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
