---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2026.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2026.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2026.csv")

```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

Some people are getting paid nearly 200 thousand dollars, while some are getting paid much less, closer to 50 thousand. There also appear to be salary brackets, showing groups of people with the same salary. There are some people that have salaries of zero dollars, which will make the data more difficult to calculate because it will not show up as a value. 

ANSWER HERE

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
wh_salaries |>
  summarize(
    mean_salary = mean(SALARY),
    median_salary = median(SALARY)
  )
```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
wh_salaries |>
  summarize(
    mean_salary = mean(SALARY),
    median_salary = median(SALARY),
    min_salary = min(SALARY),
    max_salary = max(SALARY)
  )
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

The analysis tells me that the average salary of White House employees is over 110,000 dollars. Additionally, since the mean is also near 110 thousand, it tells me that the data is pretty evenly distributed. But since the numbers are so high to begin with, the average salary is much higher than a regular constituent of the US population. There are some outliers, such as the people that make 0, as well as the people that make under 60 thousand. But for the most part, the data is relatively distributed.

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

My best story idea is comparing the average White House salaries to the general US population. I would also investigate further why the people getting paid 0 are getting paid 0, as well as why the people getting paid the most are getting paid the most. To do this, I would look further at their position description and try to access records to see if they have any special connection to the White House.
