---
title: "HW4 MD CityCrime"
---

In this assignment, we will be working with a dataset of total crimes reported by Maryland cities for two consecutive years. You'll load the data, calculate the violent and property crime rates for each city for both years, and then calculate the percentage change in the crime rates between the two years. You'll also write a short story based on your findings.

### Task 1: Load the Tidyverse and Janitor

You will need to load the tidyverse library and the janitor library

```{r}
library(tidyverse)
library(janitor)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv. You will create a dataframe called `md_cities` that contains the data from this URL.

### Task 2: Load the Data

```{r}
md_cities <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv")
```
### Task 3: Calculating Rates

Calculate violent and property crime rates for each city for both years using `mutate()`. You'll create four new columns: `violent_rate_2019`, `property_rate_2019`, `violent_rate_2020`, and `property_rate_2020`. You'll need to figure out the proper unit rate, given that there are varying populations. There's no perfect choice; choose the one that works best for you. Save the results to a new dataframe called `md_cities_rates`.

```{r}
md_cities_rates <- md_cities |>
  mutate(
  violent_rate_2019 = (violent2019 / pop2019) * 1000,
  violent_rate_2020 = (violent2020 / pop2020) * 1000,
  property_rate_2019 = (property2019 / pop2019) * 1000,
  property_rate_2020 = (property2020 / pop2020) * 1000
  )
```

### Task 4: Describe the Rates

Examine the new dataframe, `md_cities_rates`, by clicking on it in the Environment tab in RStudio and sorting by the rate columns. Write a couple of sentences about the results, comparing and contrasting the cities with the highest rates (both violent and property) in 2019 and 2020. Do you see any patterns? What does this tell you about crime in Maryland cities? What's the story here?

Ocean City, MD has some of the highest rates across the board, leading all cities in 3 of the 4 categories despite a population under 7000. Baltimore, despite being the city that statewide seems to be most known for crime, is not even in the top 10 cities in either 2020 category, a change from being #2 or #1 in 2019 in the two categories. In addition, violent crime rates decreased in 2020 from 2019, with the highest numbers in 2020 being about 2-3 crimes lower than the highest in 2019.


### Task 5: Calculate Percentage Change

Let's add two percentage change columns to your dataframe: one for violent crime rates and one for property crime rates. You'll calculate the percentage change between 2019 and 2020 for each city. Save the results to a new dataframe called `md_cities_change`.

```{r}
md_cities_change <- md_cities_rates |>
  mutate (
    violent_pct_change = (violent2020-violent2019) / (violent2019) * 100,
    property_pct_change = (property2020-property2019) / (property2019) * 100
  )
```

### Task 7: Describe Percentage Changes

Look at the `md_cities_change` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change results. What's more interesting here, the patterns or the outliers? What's the story in the data?

It seems to me that property crime has seen a general decrease across the state, while violent crime has seen a slight increase across the state. To me, the patterns are more interesting because the outliers often have a very small sample size (i.e., increasing from 1 to 3 of a certain crime will skew the data). I'd write a story comparing the rise in violent crimes to an anticipated lowering in regular crimes (due to the COVID-19 pandemic, I'd expect crime to be generally down across the state). I'd need to verify that, but I'd definitely be looking at the increase in violent crimes in many towns, with some of them very sharp increases, like Ocean City.

### Task 8: Tell the Best Story

Write up a lede and nutgraf of a story with the _best_ story idea you've found in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

Over 60% of surveyed cities report an increase in violent crime rates from 2019 to 2020, despite a general decrease in property crimes, data shows.

Even though Maryland state crime rates decreased by 15% from 2019 to 2020, according to the Maryland Statistical Analysis Center, many cities show a significant increase in their violent crime rates. Some of the most notable increases come on Maryland's Eastern Shore, in cities like Ocean City, Denton, and Easton. 

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
