---
title: "HW8_MandMs"
---

In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv

```{r}
mandm <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv")
```

### Task 3: Sample 5 rows of data from the dataset
```{r}
sample_5 <- mandm |> sample_n(5)
```

### Task 4: Calculate the mean age and standard deviation of the sample, using either the red or black number of M&Ms

```{r}
sample_5 |> summarize(mean = mean(black), sd = sd(black))
```

### Task 5: Calculate the mean age and standard deviation of the population, using the same measure as in Task 4

```{r}
mandm |> summarize(mean = mean (black), sd = sd(black))
```

### Task 6: Compare the sample and population statistics

Compare the two sets of means and standard deviations. Are they similar?

They are relatively similar, but both figures are higher in the population sample than in the sample of 5. The mean of the population is about 5.5 higher than the mean of the sample. This means that a larger sample is usually more accurate and more trustworthy.


### Task 7: Describe the results

Write a few sentences about what you found in the data. What does this tell you about the M&Ms dataset and about the act of random sampling?

I found that the data shows that the M&Ms dataset sample has vastly different numbers than the entire dataset. This tells me that the sample size needs to be big enough to accurately represent the population. If the sample is too small, it really won't demonstrate the actual population. In order for the data to be more accurate, you need to have a larger population, and random sampling is okay as long as you take a large sample.


When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
